Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2014-06-13 Referee2_Response.tex

\bigskip

\noi  
{\large \bf Specific Comments:}
\medskip

\begin{itemize}
\item[] \textit{Page 1, Line 32 (and elsewhere): I believe that you ``apply the OBM method to stochastic programming'' rather than extend it.}
\end{itemize}

\noi
Changed.  
We also went through the paper to correct other occurrences. 
In particular, this was also used in Section 6 (Conclusions). 
We changed the first sentence of Section 6 (Conclusions) to use the word `apply' instead of `extend'.

\medskip

@@

\begin{itemize}
\item[] \textit{1, 35 (and elsewhere): ``We give conditions'' instead of ``We show conditions''}
\end{itemize}

\noi
Changed. 
We also went through the paper and corrected other occurrences. 
The current version of the paper either has `give conditions' or `provide conditions'. 
\medskip

@@

\begin{itemize}
\item[] \textit{1, 37: What are the classical results? Reference?}
\end{itemize}

\noi
We reworded this sentence to improve clarity and provided references.   
\medskip

@@

\begin{itemize}
\item[] \textit{1, 49: Suggested rewrite to reduce ambiguity: ``The overlapping batch means (OBM) \ldots to obtain variance estimators having variances that are comparatively lower than their nonoverlapped counterparts. 
In the context of steady-state simulation, OBM is often used to estimate the variance of the sample mean, which is itself an estimator for the mean. 
Schmeiser et al.\ [1990] use overlapping estimators to estimate the
variance of estimators for non-means. 
More recently, overlapping estimators based on standardized time series (as opposed to batch means) have been evaluated [Alexopoulos et al., 2007a,b].'' 
In this rewrite, note the use of active and passive referencing formats.}
\end{itemize}

\noi
Thank you for this suggestion of rewrite and for your careful and detailed review of the paper. 
We rewrote this paragraph based on your corrections.
We also went through the paper and made changes to the language, paying specific attention to active and passive referencing formats. 
\medskip

@@

\begin{itemize}
\item[] \textit{4, 54 (and elsewhere): When you refer to ``variable amounts of overlap'', that is really nothing new. Certainly, any overlapping estimator will have end effects in which the overlap amounts differ from those in the ``middle'' region of the data.}
\end{itemize}

\noi
You are right.
We meant to say that the degree of overlap is parametrized and this parameter can be chosen by the user. 
We reworded this sentence to reflect this. 
xxxx  We also \ldots (I think we should write more about our conclusions and our recommendation based on our new findings here) xxxx
\medskip

\begin{itemize}
\item[] \textit{4, 57 (and elsewhere): ``classical case'' refers to nonoverlapping batch means, right?}
\end{itemize}

\noi
No; by `classical case', we meant the use of overlapping batches in simulation output analysis (as opposed to the assessment of solution quality in stochastic programming). 
We reworded this sentence to make it clear. 
Please note that based on your earlier comments, we also reworded this sentence to avoid using the term `variance reduction'. 
We also went through the whole paper and removed several uses of the word `classical' to reduce ambiguity (as these were referring to nonoverlapping batches). 
\medskip

@@

\begin{itemize}
\item[] \textit{15, 25: Do you discuss $t_{d_l,\alpha}$ at all in Theorem 3?}
\end{itemize}

\noi
No, we did not.   
We added explanations in the proof Theorem 3, discussing how $t_{d_l,\alpha}$ factors in the proof. 
\medskip

\begin{itemize}
\item[] \textit{18: Your figures are very difficult to read, even with the pdf file.}
\end{itemize}

\noi
You are right. 
We split these two figures into three and enlarged each graph. 
All the new graphs added to the paper in Section 5.4 (Computational Efficiency) are also larger than the old graphs.  
\medskip

### 2014-06-13 omrp_lb_rev.tex
A finer pass through Ref2 comments - PART I

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonoverlapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the classical nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an overlapping batches version of MRP, allowing for partial overlap, and provide conditions under which the resulting gap estimators exhibit reduced variance when compared to their counterparts obtained via nonoverlapping batches.
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We also present computational results to examine small-sample behavior. 
Our experiments indicate that the use of overlapping estimators results in gap estimators achieving lower variance than their nonoverlapping counterparts, along with comparable bias and interval coverage---for both small- and large-sample cases.

@@

Our aim is to apply the idea of overlapping batches to MRP.  
Overlapping could help to obtain better variance estimators with simple data reuse, resulting in improved CIs.
%Especially for problems where sampling error is dominant, this simple data reuse method could yield better estimators.
We note several differences in this setting compared to the classical setting.  
In simulation output analysis, the point of interest is that of estimating the variance of the sample mean of a covariance stationary process.  
We are interested in estimating the variance of an optimality gap estimator.  
Notice that an optimality gap estimator not only has a sample mean $(\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti))$ but also a minimized sample mean $(\zs_m)$.  
Minimization changes the statistical properties of sample means.  
For instance, the central limit theorem may not hold for a minimized sample mean, even though it holds for each $\x \in \X$.  
Therefore, analysis of the asymptotic benefits of overlapping becomes more complicated.
We overcome this technical difficulty by approximating the optimality gap estimators by their nonoptimized counterparts (see \S \ref{sec:theory}).  
The nonoptimized counterparts are (regular) overlapping batch means estimators (without any optimization present); therefore, they have the desired statistical properties of approximately the same bias and lower variances compared to their nonoverlapping versions. 
We establish convergence of the optimality gap estimators that use overlapping to their nonoptimized counterparts. 
Another difference between our setting and simulation output analysis is that once the data are generated through a simulation, they can be reused without much additional computational effort to obtain the OBM variance estimator.  
In this setting, due to the need to solve a sampling problem (SP$_m$), the computational effort may increase with data reuse.  
This can be alleviated in two ways. 
First, near-optimal variance reduction can be obtained by partially overlapping the batches \citep{Welch1987,Song1992}.  
Partial overlapping results in a smaller number of batches; hence, fewer optimization problems need to be solved.
In this paper, we consider variable amounts of overlap and examine this both theoretically and empirically. 
Second, in many solution methods, warm-starting can be used to quickly solve the sampling approximations with overlapping samples, considerably reducing solution time.

@@

As before, let $m$ denote the batch size, $n$ the total sample size and $k = \lfloor \tfrac{n}{m} \rfloor$ be the number of nonoverlapping batches.  
In this paper, we use the {\it batch nonoverlap parameter} $1 \leq \gamma \leq m$ to denote how much neighboring batches do {\it not} overlap.  
For instance, $\gamma = m$ corresponds to the classical case of nonoverlapping batches and $\gamma = 1$ corresponds to the maximally overlapping case of \citet{Meketon1984}.  
In general, $\gamma$ can take any integer value between and including $[1,m]$. 
Let $\lfloor \cdot \rfloor$ be the greatest integer that is less than or equal to its argument. 
The sample mean of each batch estimator is calculated similarly, 
$
\yb_j = \frac{1}{m} \sum_{i=1}^m y^{\gamma(j-1) + i},\ \  j = 1, 2, \dots, \nb, 
$
where $\nb$ is the number of batches used given $n$, $m$ and $\gamma$.  
The sample variance estimator given in (\ref{eq:overlap_variance_def}) is changed to
\begin{equation} \label{eq:variance_gamma}
	\vartg{\ybb} = \frac{1}{\left( \tfrac{n}{m} - 1 \right) \left( \nb \right)}  \sum_{j=1}^{\nb} (\yb_j - \ybb)^2.
\end{equation}
Note that with $\gamma=1$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:overlap_variance_def}) and with $\gamma=m$ and $n=mk$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:var}).

@@

\begin{description}
	\item[A1] Samples of the random variable $\xit$ are i.i.d.
	\item[A2] $\sup_{\x \in X} \e{f(\x,\xi)^{4}} < \infty$.
	\item[A3] (\ref{eq:sto_prog}) has a unique optimal solution $\xs$ and $\p{\xs_m \neq \xs} \rightarrow 0$ exponentially fast as $m \rightarrow \infty$.  
           That is, there exists constants $M > 0$ and $c > 0$ such that $\log(\p{\xs_m \neq \xs}) \leq -cm$ for all $m > M$.
\end{description}

First, we restrict our attention to i.i.d.\ sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of the sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied for different classes of problems with unique optimal solutions.  
\citet{shapiro2000rate} establish exponential rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite.   
\citet{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

Theorem \ref{thm:varvar_conv} shows conditions under which the same asymptotic variance reduction results are achieved for the OMRP optimality gap variance estimator. 
We will empirically examine the variance reduction for small sample sizes in \S \ref{sec:comp}. 
%We end our theoretical properties in the next section by examining the CIs formed using OMRP.

@@

Finally, Figure~\ref{fig:nv}(b) shows the coverage probability of CIs generated by OMRP for several values of $m$ across varying values of $\gamma/m$ for the newsvendor problem.  
The results of the classical overlapping batches estimators show that coverage probability does not change with the amount of overlap, which we can  empirically see in this figure for OMRP for NVD.
We observed the same for CEP1, 10D, and DB1.  
The coverage probabilities from applying the (nonoverlapping) MRP algorithm to these problems presented in \citep{Bayraksan2006} agree with our results: for the newsvendor problem, coverage probability drops as $m$ increases and the coverage probability of CEP1 (not shown) remains fairly constant around the desired value of $90\%$. 
Coverage probability for 10D (not shown) and {\it estimated} coverage probability for DB1 (also not shown) were very high, more than $0.99$.  
The bias from solving (\ref{eq:sto_prog_m}) for these problems seems to be much more significant than the variance.

@@

We have extended previous work from simulation output analysis to assessing solution quality in stochastic programming by means of variably overlapping batches \citep{Meketon1984,Song1992,Welch1987} for Monte Carlo simulation-based estimators of optimality gaps in stochastic programs \citep{Mak1999}. 
We have provided conditions under which the resulting point estimator of the optimality gap and its associated sample variance are consistent, the same variance reduction can be achieved, and asymptotically valid CIs can be formed. 
Empirical results with small sample sizes indicate that the asymptotic reductions in variance of the variance estimator show a similar decrease in OMRP, while bias and coverage probability remain unaffected.

>>>>

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonoverlapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an overlapping batches version of MRP, allowing for partial overlap, and provide conditions under which the resulting gap estimators exhibit reduced variance when compared to their counterparts obtained via nonoverlapping batches.
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We also present computational results to examine small-sample behavior. 
Our experiments indicate that the use of overlapping estimators results in gap estimators achieving lower variance than their nonoverlapping counterparts, along with comparable bias and interval coverage---for both small- and large-sample cases.

@@

Our aim is to apply the idea of overlapping batches to MRP.  
Overlapping could help to obtain better variance estimators with simple data reuse, resulting in improved CIs.
%Especially for problems where sampling error is dominant, this simple data reuse method could yield better estimators.
We note several differences in this setting compared to the classical simulation setting.  
In simulation output analysis, the point of interest is that of estimating the variance of the sample mean of a covariance stationary process.  
We are interested in estimating the variance of an optimality gap estimator.  
Notice that an optimality gap estimator not only has a sample mean $(\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti))$ but also a minimized sample mean $(\zs_m)$.  
Minimization changes the statistical properties of sample means.  
For instance, the central limit theorem may not hold for a minimized sample mean, even though it holds for each $\x \in \X$.  
Therefore, analysis of the asymptotic benefits of overlapping becomes more complicated.
We overcome this technical difficulty by approximating the optimality gap estimators by their nonoptimized counterparts (see \S \ref{sec:theory}).  
The nonoptimized counterparts are (regular) overlapping batch means estimators (without any optimization present); therefore, they have the desired statistical properties of approximately the same bias and lower variances compared to their nonoverlapping versions. 
We establish convergence of the optimality gap estimators that use overlapping to their nonoptimized counterparts. 
Another difference between our setting and simulation output analysis is that once the data are generated through a simulation, they can be reused without much additional computational effort to obtain the OBM variance estimator.  
In this setting, due to the need to solve a sampling problem (SP$_m$), the computational effort may increase with data reuse.  
This can be alleviated in two ways. 
First, near-optimal variance reduction can be obtained by partially overlapping the batches \citep{Welch1987,Song1992}.  
Partial overlapping results in a smaller number of batches; hence, fewer optimization problems need to be solved.
In this paper, we consider variable amounts of overlap and examine this both theoretically and empirically. 
Second, in many solution methods, warm-starting can be used to quickly solve the sampling approximations with overlapping samples, considerably reducing solution time.

@@

As before, let $m$ denote the batch size, $n$ the total sample size and $k = \lfloor \tfrac{n}{m} \rfloor$ be the number of nonoverlapping batches.  
In this paper, we use the {\it batch nonoverlap parameter} $1 \leq \gamma \leq m$ to denote how much neighboring batches do {\it not} overlap.  
For instance, $\gamma = m$ corresponds to the case of nonoverlapping batches and $\gamma = 1$ corresponds to the maximally overlapping case of \citet{Meketon1984}.  
In general, $\gamma$ can take any integer value between and including $[1,m]$. 
Let $\lfloor \cdot \rfloor$ be the greatest integer that is less than or equal to its argument. 
The sample mean of each batch estimator is calculated similarly, 
$
\yb_j = \frac{1}{m} \sum_{i=1}^m y^{\gamma(j-1) + i},\ \  j = 1, 2, \dots, \nb, 
$
where $\nb$ is the number of batches used given $n$, $m$ and $\gamma$.  
The sample variance estimator given in (\ref{eq:overlap_variance_def}) is changed to
\begin{equation} \label{eq:variance_gamma}
	\vartg{\ybb} = \frac{1}{\left( \tfrac{n}{m} - 1 \right) \left( \nb \right)}  \sum_{j=1}^{\nb} (\yb_j - \ybb)^2.
\end{equation}
Note that with $\gamma=1$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:overlap_variance_def}) and with $\gamma=m$ and $n=mk$, (\ref{eq:variance_gamma}) reduces to (\ref{eq:var}).

@@

\begin{description}
	\item[A1] Samples of the random variable $\xit$ are i.i.d.
	\item[A2] $\sup_{\x \in X} \e{f(\x,\xit)^{4}} < \infty$.
	\item[A3] (\ref{eq:sto_prog}) has a unique optimal solution $\xs$ and $\p{\xs_m \neq \xs} \rightarrow 0$ exponentially fast as $m \rightarrow \infty$.  
           That is, there exists constants $M > 0$ and $c > 0$ such that $\log(\p{\xs_m \neq \xs}) \leq -cm$ for all $m > M$.
\end{description}

First, we restrict our attention to i.i.d.\ sampling by A1.
The results for overlapping batches from the simulation output analysis indicate that $\var{n_l VG_l}$ may be reduced by overlapping, thus requiring at least finite fourth moments. 
We will formally show this result in Section~\ref{ssec:var_reduct}.
Our proof relies on an exponential rate of convergence of the sampled solution to the true solution of the problem, specified by A3.  
Assumption A3 can be satisfied for different classes of problems with unique optimal solutions.  
\citet{shapiro2000rate} establish exponential rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite.   
\citet{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $\x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

Theorem \ref{thm:varvar_conv} gives conditions under which the same asymptotic variance reduction results are achieved for the OMRP optimality gap variance estimator. 
We will empirically examine the variance reduction for small sample sizes in \S \ref{sec:comp}. 
%We end our theoretical properties in the next section by examining the CIs formed using OMRP.

@@

Finally, Figure~\ref{fig:nv}(b) shows the coverage probability of CIs generated by OMRP for several values of $m$ across varying values of $\gamma/m$ for the newsvendor problem.  
The results of the overlapping batches estimators from the simulation literature show that coverage probability does not change with the amount of overlap, which we can  empirically see in this figure for OMRP for NVD.
We observed the same for CEP1, 10D, and DB1.  
The coverage probabilities from applying the (nonoverlapping) MRP algorithm to these problems presented in \citep{Bayraksan2006} agree with our results: for the newsvendor problem, coverage probability drops as $m$ increases and the coverage probability of CEP1 (not shown for brevity) remains fairly constant around the desired value of $90\%$. 
Coverage probability for 10D (not shown) and {\it estimated} coverage probability for DB1 (also not shown) were very high, more than $0.99$.  
The bias from solving (\ref{eq:sto_prog_m}) for these problems seems to be much more significant than the variance.

@@

We have applied the overlapping batch means method from simulation output analysis \citep{Meketon1984,Song1992,Welch1987} to assessing solution quality in stochastic programming, in particular, to Monte Carlo simulation-based estimators of optimality gaps \citep{Mak1999}. 
We have provided conditions under which the resulting point estimator of the optimality gap and its associated sample variance are consistent, the same variance reduction can be achieved, and asymptotically valid CIs can be formed. 
Empirical results with small sample sizes indicate that the asymptotic reductions in variance of the variance estimator show a similar decrease in OMRP, while bias and coverage probability remain unaffected.
